---
title: "Lecture Notes for Statistics for Biology FS2020"
author: "Stephan Peischl"
date: "1/10/2020"

output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

These notes should help you get started with the statistics software R, guide you through the practical part of the course and provide additional information on selected topics. Before you continue reading, make sure you have R and R Studio installed on your computer. Try to use R as an calcualtor as shown in the lecture slides and try to get yourelf acustomed to the different windows in the RStudio envrionment. No prior knwoledge about R is necessary and once everything is up and running yu can dive right in.

We start by loading the relevant packages for this course. An R package is a collection of functions, data, and documentation that extends the capabilities of base R.  We are going to make use of the so-called tidyverse. 

```{r libraries.eval,echo = TRUE,results='hide', message=F, warning=F}
# make sure you have th packages installed using 
# install.packages(packagename)

library(tidyverse)
library(Rmisc)
```



# Visualization using ggplot

We are going start with some examples of how to visulaize data. This should show you how easy it is to produce high quality visulaizations using R and motivate you to learn more about programming and data analysis with R. You will be learning the basic structure of a ggplot2 plot, and powerful techniques for turning data into plots.

R has several systems for making graphs, and I have chosen to start with ggplot2 beacuse it is elegant, versatile and is easy to leanr if you have no prior knowledge about programming. For those of you who are already familiar wiht programming langauges, doing graphics in "base R" may seem more intuitive initially but I hope I can convince you that the grammar of plotting used by ggplot has lots of advantages and is very readable once you get the hang of it. 

If you’d like to learn more about the theoretical underpinnings of ggplot2 before you start, I’d recommend reading “The Layered Grammar of Graphics”, http://vita.had.co.nz/papers/layered-grammar.pdf.

### The MPG dataset
Let’s use our first graph to answer a question: Do cars with big engines use more fuel than cars with small engines? What does the relationship between engine size and fuel efficiency look like? Is it positive? Negative? Linear? Nonlinear?

We can find an answer with the mpg data set found in ggplot2 (aka ggplot2::mpg). A data frame is the R data format to store data in the format of a spreadsheet: rectangular collection of variables (in the columns) and observations (in the rows). mpg contains observations collected by the US Environmental Protection Agency on 38 models of car.

```{r dataframe}
mpg
```

Among the variables in mpg are:



**displ**, a car’s engine size, in litres.

**hwy**, a car’s fuel efficiency on the highway, in miles per gallon (mpg). A car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance.



To learn more about mpg, open its help page by running ?mpg. in the next seciton we learn how to visualize such data.




### The grammar of ggplot2
In general you begin a plot with the function *ggplot()*. ggplot() creates a coordinate system that you can add layers to. The first argument of ggplot() is the dataset to use in the graph. For instance, typing 

```{r, echo = T,results='hide', message=F, warning=F}
ggplot(data = mpg) 
```

into the console creates an empty graph (which is boring so it is not shown here). You can now add layers to your plot. For example, the function geom_point() adds a layer of points to your plot creating a a scatterplot.

```{r mileage}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) 
```

There are many other such *geometirc functions" withing the ggplot2 universe and you can combine them in a single plot.  You’ll learn a whole bunch of them throughout this course, but be aware that we will only scratch the surface of what is available. After this course, you should be equipped with everthing you need to equip your self with new tools for data analysis whenever you need them.

Each geom function in ggplot2 takes a mapping argument. This defines how variables in your dataset are mapped to visual properties. The mapping argument is always paired with aes(), and the x and y arguments of aes() specify which variables to map to the x and y axes. ggplot2 looks for the mapped variables in the data argument, in this case, mpg.

Next, we use a different theme beacuse this gray background in the standard design is ugly.

```{r mileage2}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  theme_classic()
```

The general grammar of ggplot is

ggplot(data = \<DATA\>) + 

 \<GEOM_FUNCTION\>(mapping = aes(\<MAPPINGS\>))
  
We will explain each part step by step in examples. The DATA argument should be quite clear: here you specify the dataframe (we will learn more about that alter) that you want to work with.

The second part, GEOM_FUNCTION, specifies what kind of plot you want. for instance, geom_point gives you a classic scatter plot, whereas geom_smooth would give you a smooth line theat best resembles the cloud of points (more on  that later):

```{r mileage3}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) +
  theme_classic()
```

Of course, we can put points and a smooth line on the same plot: 
```{r mileage4}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  theme_classic()
```

You may wonder how the smooth line is determined. We will come back to this later in the chapter on linear regression. For now, we just mention that we can modify the method used for the plot, e.g., to get a linear regression line:
```{r mileage_lm}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy),method='lm') +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  theme_classic()
```

Finally, MAPPING specifies which variables hould be plotted. We ussually specify this with an aesthetics mapping of the form mapping = aes(x = ..., y =...) to deterimne the variables that give us the x- and y-coordinates. We can however also change the aesthetics of our plot in various way, for instance by giving data points different colors that indicate the value of a third variable:  
```{r mileage_aes}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy,color = class)) +
  theme_classic()
```

Another important option is to set a clor manually:
```{r mileage_aes_2}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") +
  theme_classic()
```

As a final remark, you can also create a plot, store it in a variable and then explore themes after that:

```{r mileage_aes_themes}
p = ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") 

p+theme_classic()
p+theme_dark()
```

# Descriptive Statistics 

## Calculating statiscis in R

## Visulizing distirbutions 
We can visualize continous data in a histogram, for instnace how many cars we have in each class:

```{r}
ggplot(data = mpg) + 
  geom_bar(mapping = aes(x = class))
```

We can make stacked graphs using the argument $fill$ and a discrete variable such as number of cylinders 

```{r}
ggplot(data = mpg) + 
  geom_bar(mapping = aes(x = class, fill = as.factor(cyl)))
```

If we want to show the distribution of a continous variable for each class, we can do boxplots, e.g., showing the mileage of each class
```{r}
ggplot(data = mpg) + 
geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) + 
coord_flip()
```



Another way of summarizing data is by showing several statistics using whisker plots. A simple solution is using the stat_summary function. In the example below the dots show the median, and the whiskers extend to the minium and maximum in each category (class).

```{r}
ggplot(data = mpg) + 
  stat_summary(
    mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  ) +
  coord_flip()
```

If we want to show confidnece intervalls (or any otehr quantity), a clean way to do this is to simply create a data set containing this information and then use it to plot the quantitites.

```{r}

mpgSE <- summarySE(mpg, measurevar="hwy", groupvars=c("class"))

mpgSE

ggplot(mpgSE, aes(x = reorder(class, hwy, FUN = median), y = hwy)) + 
    geom_errorbar(aes(ymin=hwy-ci, ymax=hwy+ci), width=.1) +
    geom_point() +
    coord_flip()
```



# Loading datastes

# A bit of workflow

# Distributions

## Binomial 

## Normal 

# Hypothesis Tests

## Binomial-test, t-test

# Deviations from Normailty

## Transformations

## Non-Parametric Alterantives

## Permutation Test

# Testing for independence

# ANOVA

# Linear Regression